{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce

### 2013-07-31 manuscript/bmc_additional1.tex
Manuscript in BMC format

\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{url}
\usepackage{multirow}
\usepackage{authblk}
\usepackage{rotating}
\usepackage{xr}

% \renewcommand{\thetable}{S\arabic{table}}   
% \renewcommand{\thefigure}{S\arabic{figure}}
% \renewcommand{\thealgorithm}{\arabic{algorithm}}

\usepackage[style=nature,maxbibnames=5,mincitenames=1,maxcitenames=1]{biblatex}
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
\addbibresource{manuscript.bib}

\renewcommand{\figurename}{Supplementary Figure}
\renewcommand{\tablename}{Supplementary Table}

\newfloat{suppalgorithm}{p}{cap}
\floatname{suppalgorithm}{Supplementary Algorithm}

\begin{document}

\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce \\
Supplementary Material}

\author[1,5]{Christopher W. Whelan \thanks{whelanch@ohsu.edu}}
\author[3,4]{Jeffrey Tyner}
\author[6]{Alberto L'Abbate}
\author[6]{Clelia Tiziana Storlazzi}
\author[4,5]{Lucia Carbone}
\author[1,2,5]{Kemal S\"onmez \thanks{sonmezk@ohsu.edu}}
\affil[1]{Institute on Development and Disability and Center for Spoken Language Understanding}
\affil[2]{Department of Medical Informatics \& Clinical Epidemiology}
\affil[3]{Program in Molecular and Cellular Biosciences}
\affil[4]{Behavioral Neuroscience Department}
\affil[5]{Oregon Health \& Science University, Portland, OR, USA}
\affil[6]{Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy}

@@

\begin{suppalgorithm}[h]
\algrenewcommand\algorithmicprocedure{\textbf{job}}
  \begin{algorithmic}[1]
    \Procedure{Alignment}{}
    \Function{Map}{$\textrm{ReadPairId }rpid, \textrm{ReadId }r, \textrm{ReadSequence }s, \textrm{ReadQuality }q$}
    \ForAll{$ \textrm{Alignments }a \in \textsc{Align}(<s,q>)$}
    \State $\textsc{Emit}(\textrm{ReadPairId }rpid, \textrm{Alignment }a)$
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{ReadPairId }rpid, \textrm{Alignments }a_{1,2,\ldots}$}
    \State $\textrm{AlignmentPairList }ap \gets \textsc{ValidAlignmentPairs}(a_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{ReadPairId }rp, \textrm{AlignmentPairList } ap)$
    \EndFunction
    \EndProcedure

    \Procedure{Compute SV Features}{}
    \Function{Map}{$\textrm{ReadPairId }rp, \textrm{AlignmentPairList }ap$}
    \ForAll{$ \textrm{AlignmentPairs }<a_1,a_2>  \in ap$}
    \ForAll{$ \textrm{GenomicLocations }l \in \textsc{Loci }(a_1,a_2)$}
    \State $ \textrm{ReadPairInfo }rpi \gets <\textrm{InsertSize}(a_1,a_2), \textrm{AlignmentScore}(a_1,a_2)>$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{ReadPairInfo }rpi)$
    \EndFor
    \EndFor
    \EndFunction
    \Function{Reduce}{$\textrm{GenomicLocation }l, \textrm{ReadPairInfos }rpi_{1,2,\ldots}$}
    \State $\textrm{SVFeatures } \phi_l \gets \Phi(\textrm{InsertSizes }i_{1,2,\ldots}, \textrm{AlignmentScores }q_{1,2,\ldots})$
    \State $\textsc{Emit}(\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l)$
    \EndFunction
    \EndProcedure

    \Procedure{Call SVs}{}
    \Function{Map}{$\textrm{GenomicLocation }l, \textrm{SVFeatures } \phi_l$}
    \State $\textsc{Emit}(\textrm{Chromosome}(l), <l,\phi_l>)$
    \EndFunction
    \Function{Reduce}{$\textrm{Chromosome }c, \textrm{GenomicLocation } l_{1,2,\ldots},\phi_{1,2,\ldots}$}
    \State $\textrm{StructuralVariationCalls } svs_c \gets \textsc{PostProcess }(\phi_{1,2,\ldots})$
    \EndFunction
    \EndProcedure
  \end{algorithmic}
\caption{The algorithmic framework for SV calling in MapReduce.}
\label{cb_algo}
\end{suppalgorithm}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{rrrrrr}
  \hline
 & 40-100bp  & 101-250bp  & 251-500bp & 501-1000bp & $>$ 1000bp \\ 
 Total Number & 224 &  84 & 82 &  31 & 26\\ 
  \hline
  Cloudbreak  & \textbf{68} (17)  & \textbf{67} (\textbf{10}) &  \textbf{56} (\textbf{5}) & \textbf{11} (\textbf{3}) & \textbf{15} (\textbf{0}) \\ 
  Breakdancer & 52 (8)  & 49 (2) &  49 (0) & 7 (0) & 14 (\textbf{0}) \\ 
  GASVPro     & 35 (2)  & 26 (0) &  26 (0) & 2 (0) & 6 (\textbf{0}) \\ 
  DELLY-RP       & 22 (1)  & 56 (1) &  40 (0) & 8 (0) & 12 (\textbf{0}) \\ 
  DELLY-SR       & 0 (0)  & 2 (0) &  28 (0) & 2 (0) & 10 (\textbf{0}) \\ 
  Pindel      & 60 (\textbf{32})  & 16 (0) &  41 (2) & 1 (0) & 12 (\textbf{0})\\ 
   \hline
\end{tabular}
\end{center}
\caption{The number of simulated deletions in the 30X diploid chromosome 2 with Venter indels found by each tool at a 10\% FDR, as well as the number of those deletions that were discovered exclusively by each tool (in parentheses). The total number of deletions in each size class in the true set of deletions is shown in the second row of the header.}
\label{chr2DeletionPredsFDR10}
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|rr|rr|}
\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\cline{3-6}
\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
\cline{2-6}
\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 35 & 2 &  96 & 21 \\
 & Heterozygous & 0 & 39 &  2 & 448 \\
\cline{2-6}
\end{tabular}
\end{center}
\caption{Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
\label{deletionGenotypeaccuracy}
\end{table}

@@

\section{Supplementary Discussion}

\subsection{A general framework for SV Detection in MapReduce}

We have developed a conceptual algorithmic framework for SV detection in MapReduce, which is outlined in Supplementary Algorithm \ref{cb_algo}. As described in the main text, this framework divides processing into three separate MapReduce jobs: an alignment job, a feature computation job, and an SV calling job.

The alignment job uses sensitive mapping tools and settings to discover mapping locations for each read pair. Aligners can be executed to report multiple possible mappings for each read, or only the best possible mapping. Given a set of read pairs, each of which consists of a read pair identifier $rpid$ and two sets of sequence and quality scores $<s,q>$, each mapper aligns each pair end set $<s,q>$ in either single- or paired end mode and emits possible mapping locations under the $rpid$ key. Reducers then collect the alignments for each paired end, making them available under one key for the next job. Our implementation contains wrappers to execute the aligners BWA, GEM \autocite{MarcoSola:2012hm}, Novoalign\footnote{\url{http://wwww.novocraft.com}}, RazerS 3 \autocite{Weese:2012by}, mrFAST \autocite{Alkan:2009cr}, and Bowtie 2 \autocite{Langmead:2012jh}, as well as the ability to import a pre-aligned BAM file directly to HDFS.

In the second job, we compute a set of features for each location in the genome. To begin, we tile the genome with small fixed-width, non-overlapping intervals. For the experiments reported in this paper we use an interval size of 25bp. Let $L = \left\{l_1,l_2,\ldots,l_N\right\}$ be the set of intervals covering the entire genome. Let $R^1 = \left\{r^{1}_{1},r^{1}_{2},\ldots,r^{1}_{M}\right\}$ and $R^2 = \left\{r^{2}_{1},r^{2}_{2},\ldots,r^{2}_{M}\right\}$ be the input set of paired reads. Let $A^1 = \left\{a^{1}_{m,1},a^{1}_{m,2},\ldots,a^{1}_{m,K}\right\}$ and $A^2 = \left\{a^{2}_{m,1},a^{2}_{m,2},\ldots,a^{2}_{m,L}\right\}$ be the set of alignments for the left and right reads from read pair $m$. For any given pair of alignments of the two reads in a read pair, $a^{1}_{m,i}$ and $a^{2}_{m,j}$, let the $\textrm{ReadPairInfo } rpi_{m,i,j}$ be information about the pair relevant to detecting SVs, e.g. the fragment size implied by the alignments and the likelihood the alignments are correct. We then leave two functions to be implemented depending on the application:
\begin{flalign*}
 \textsc{Loci } :& \langle a^{1}_{m,i},a^{2}_{m,j} \rangle \rightarrow L_m \subseteq L \\
 \Phi :& \left\{\textrm{ReadPairInfo }rpi_{m,i,j}\right\} \rightarrow \mathbb{R}^N \\
\end{flalign*}

The first function, \textsc{Loci}, maps an alignment pair to a set of genomic locations to which it is relevant for SV detection; for example, the set of locations overlapped by the internal insert implied by the read alignments. We optimize this step by assuming that if there exist concordant mappings for a read pair, defined as those where the two alignments are in the proper orientation and with an insert size within three standard deviations of the expected library insert size, one of them is likely to be correct and therefore we do not consider any discordant alignments of the pair. The second function, $\Phi$, maps a set of ReadPairInfos relevant to a given location to a set of real-valued vectors of features useful for SV detection.

Finally, the third MapReduce job is responsible for making SV calls based on the features computed at each genomic location. It calls another application-specific function  $\textsc{PostProcess} : \left\{\phi_1,\phi_2,\ldots,\phi_N\right\} \rightarrow \left\{\langle  \textrm{SVType } s, l_{start}, l_{end} \rangle\right\}$  that maps the sets of features for related loci into a set of SV calls characterized by their type $s$ (i.e Deletion, Insertion, etc.) and their breakpoint locations $l_{start}$ and $l_{end}$. We parallelize this job in MapReduce by making calls for each chromosome in parallel, which we achieve by associating a location and its set of features to its chromosome in the map phase, and then making SV calls for one chromosome in each reduce task.

\subsection{Cloudbreak: An Implementation of the MapReduce SV Detection Framework}

Our Cloudbreak software package can be seen as an implementation of the general framework defined above. In particular, Cloudbreak implements the three user-defined function described above as follows:

\begin{description}
\item[\sc{Loci}] Because we are detecting deletions and short insertions, we map ReadPairInfos from each possible alignment to the genomic locations overlapped by the implied internal insert between the reads. For efficiency, we define a maximum detectable deletion size of 25,000bp, and therefore alignment pairs in which the ends are more than 25kb apart, or in the incorrect orientation, map to no genomic locations.
\item[$\Phi$] To compute features for each genomic location, we follow \textcite{Lee:2009da}, who observed that if all mappings are correct, the insert sizes implied by mappings which span a given genomic location should follow a Gaussian mixture model (GMM) whose parameters depend on whether a deletion or insertion is present at that locus (Supplementary Figure \ref{insert_size_mixes}). Briefly: if there is no indel, the insert sizes implied by spanning alignment pairs should follow the distribution of actual fragment sizes in the sample, which is typically modeled as normally distributed with mean $\mu$ and standard deviation $\sigma$. If there is a homozygous deletion or insertion of length $l$ at the location, $\mu$ should be shifted to $\mu + l$, while $\sigma$ will remain constant. Finally, in the case of a heterozygous event, the distribution of insert sizes will follow a mixture of two normal distributions, one with mean $\mu$, and the other with mean $\mu + l$, both with an unchanged standard deviation of $\sigma$, and mixing parameter $\alpha$ that describes the relative weights of the two components. Because the mean and standard deviation of the fragment sizes are selected by the experimenter and therefore known \emph{a priori} (or at least easily estimated based on a sample of alignments), we only need to estimate the mean of the second component at each locus, and the mixing parameter $\alpha$.

To handle incorrect and ambiguous mappings, we assume that in general they will not form normally distributed clusters in the same way that correct mappings will, and therefore use an outlier detection technique to filter the observed insert sizes for each location. We sort the observed insert sizes and define as an outlier an observation whose $k$th nearest neighbor is more than $n\sigma$ distant, where $k = 3$ and $n = 5$. In addition, we rank all observations by the estimated probability that the mapping is correct and use an \emph{adaptive quality cutoff} to filter observations: we discard all observations where the estimated probability the mapping is correct is less than the score of the maximum quality observation minus a constant $c$. This allows the use of more uncertain mappings in repetitive regions of the genome while restricting the use of low-quality mappings in unique regions. Defining $\textsc{Mismatches}(a)$ to be the number of mismatches between a read and the reference genome in the alignment $a$, we approximate the probability $p^{k}_c$ of each end alignment being correct by:

\[ p^{k}_c(a^{k}_{m,i}) = \frac{\exp({-\textsc{Mismatches}(a^{k}_{m,i})/2)}}{\sum_j{\exp(-\textsc{Mismatches}(a^{k}_{m,j})/2)}} \]

And then multiply $p_c(a^{1}_{m,i})$ and $p_c(a^{2}_{m,i})$ to approximate the likelihood that the pair is mapped correctly.

We fit the parameters of the GMM using the Expectation-Maximization algorithm. Let $Y = y_{1,2, \ldots m}$ be the observed insert sizes at each location after filtering, and say the library has mean fragment size $\mu$ with standard deviation $\sigma$. We initialize the two components to have means $\mu$ and $\bar{Y}$, set the standard deviation of both components to $\sigma$, and set $\alpha = .5$. In the E step, we compute for each $y_i$ and GMM component $j$ the value $\gamma_{i,j}$, which is the normalized likelihood that $y_i$ was drawn from component $j$. We also compute $n_j = \sum_i{\gamma_{i,j}}$, the relative contributions of the data points to each of the two distributions. In the M step, we update $\alpha$ to be $n_2 - \left|Y\right|$, and set the mean of the second component to be $\frac{\sum_m{\gamma_{m,2}y_m}}{n_2}$. We treat the variance as fixed and do not update it, since under our assumptions the standard deviation of each component should always be $\sigma$. We repeat the E and M steps until convergence, or until a maximum number of steps has been taken.

The features generated for each location $l$ include the log-likelihood ratio of the filtered observed data points under the fit GMM to their likelihood under the distribution $N(\mu,\sigma)$, the final value of the mixing parameter $\alpha$, and $\mu'$, the estimated mean of the second GMM component.

\item[\sc{PostProcess}] We convert our features along each chromosome to insertion and deletion calls by first extracting contiguous genomic loci where the log-likelihood ratio of the two models is greater than a given threshold. To eliminate noise we apply a median filter with window size 5. We end regions when $\mu'$ changes by more than 60bp ($2\sigma$), and discard regions where the average value of $\mu'$ is less than $\mu$ or where the length of the region differs from $\mu'$ by more than $\mu$.
\end{description}

\subsection{Runtime Analysis}

@@

In Supplementary Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work. Each tool varies in the amount of parallelization supported. We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming. Pindel allows multi-threaded operation on multicore servers. Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem. Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread. DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode. GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this component. The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.

@@

\printbibliography[title=Supplementary References]

@@

\end{document}

>>>>

\documentclass[10pt]{bmc_article}

\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{url}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{xr}

\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\thetable}{A\arabic{table}}

\urlstyle{rm}

\setlength{\topmargin}{0.0cm}
\setlength{\textheight}{21.5cm}
\setlength{\oddsidemargin}{0cm} 
\setlength{\textwidth}{16.5cm}
\setlength{\columnsep}{0.6cm}

\newboolean{publ}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                              %%
%% You may change the following style settings  %%
%% Should you wish to format your article       %%
%% in a publication style for printing out and  %%
%% sharing with colleagues, but ensure that     %%
%% before submitting to BMC that the style is   %%
%% returned to the Review style setting.        %%
%%                                              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings
\newenvironment{bmcformat}{\fussy\setboolean{publ}{true}}{\fussy}

\newcommand{\degree}{\ensuremath{^\circ}}

\begin{document}
\begin{bmcformat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{{\em Cloudbreak}: Accurate and Scalable Genomic Structural Variation Detection in the Cloud with MapReduce}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the authors here                   %%
%%                                          %%
%% Ensure \and is entered between all but   %%
%% the last two authors. This will be       %%
%% replaced by a comma in the final article %%
%%                                          %%
%% Ensure there are no trailing spaces at   %% 
%% the ends of the lines                    %%     	
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\author{Christopher W Whelan\correspondingauthor$^{1}$%
       \email{Christopher W Whelan\correspondingauthor - cwhelan@gmail.com}%
      \and
         Jeffrey Tyner$^2$%
      \and
         Alberto L'Abbate$^3$%
      \and
         Clelia Tiziana Storlazzi$^3$
      \and
         Lucia Carbone$^4$
       and 
         Kemal S\"onmez\correspondingauthor$^{4,5}$%
         \email{Kemal S\"onmez\correspondingauthor - sonmezk@ohsu.edu}%
      }

\address{%
    \iid(1)Institute on Development and Disability and Center for Spoken Language Understanding, Oregon Health \& Science University, %
    3181 SW Sam Jackson Park Rd, GH40, Portland, OR 97239-3098\\
    \iid(2)Program in Molecular and Cellular Biosciences, Oregon Health \& Science University, %
    3181 SW Sam Jackson Park Rd, L102, Portland, OR 97239\\
    \iid(3)Department of Biology, University of Bari ``Aldo Moro'', Via G. Amendola 165/A, 70126, Bari, Italy\\
    \iid(4)Department of Behavioral Neuroscience, Oregon Health \& Science University, %
    3181 SW Sam Jackson Park Rd, L470, Portland, Oregon, U.S.A. 97239-3098\\
    \iid(5)Department of Medical Informatics \& Clinical Epidemiology, Oregon Health \& Science University, %
    3181 S.W. Sam Jackson Park Rd., BICC, Portland, Oregon, U.S.A. 97239-3098
}%

@@

@@

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r|r|rr|rr|}
\multicolumn{2}{c}{}  & \multicolumn{4}{c}{Actual Genotypes} \\
\multicolumn{2}{c}{}  & \multicolumn{2}{c}{Simulated Data} & \multicolumn{2}{c}{NA18507} \\
\cline{3-6}
\multicolumn{2}{c|}{} &  Homozygous & Heterozygous & Homozygous & Heterozygous \\ 
\cline{2-6}
\multirow{2}{*}{\shortstack{Predicted \\ Genotypes}} & Homozygous & 35 & 2 &  96 & 21 \\
 & Heterozygous & 0 & 39 &  2 & 448 \\
\cline{2-6}
\end{tabular}
\end{center}
\caption{Table A4. Confusion matrices for the predicted genotype of deletions found by Cloudbreak on both the simulated and NA18507 data sets.}
\label{deletionGenotypeaccuracy}
\end{table}

@@

\section{Runtime Analysis}

@@

In Table \ref{runtimes} we display a comparison of runtimes on the real and simulated data sets for all of the tools evaluated in this work. Each tool varies in the amount of parallelization supported. We report runtimes for tools run in their default single-threaded mode, as well as for levels of parallelization achievable with basic scripting, noting that one of the key advantages of Hadoop/MapReduce is the ability to scale parallel execution to the size of the available compute cluster without any custom programming. Pindel allows multi-threaded operation on multicore servers. Pindel and Breakdancer allow processing of a single chromosome in one process, so it is possible to execute all chromosomes in parallel on a cluster that has a job scheduler and shared filesystem. Breakdancer has an additional preprocessing step (\texttt{bam2cfg.pl}) which runs in a single thread. DELLY suggests splitting the input BAM file by chromosome, after which a separate DELLY process can be executed on the data for each chromosome; splitting a large BAM file is a time consuming process and consumes most of the time in this parallel workflow, in fact making it faster to run in single-threaded mode. GASVPro allows parallelization of the MCMC component for resolving ambiguously mapped read pairs; however, this requires a significant amount of custom scripting, and we did not find that the MCMC module consumed most of the runtime in our experiments, so we do not attempt to parallelize this component. The MoDIL distribution contains a set of scripts that can be used to submit parallel jobs to the SGE scheduling engine or modified for other schedulers; we adapted these for use in our cluster.

@@

@@

\end{bmcformat}
\end{document}
